VariantSpark: Population Scale Clustering of Genotype Information

### 2015-11-27 doc/publications/journal2014/paper.tex
updated times in document

As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 3 minutes. %2mins and 58secs. %178s 
This is almost 80\% faster than both the {\sc ADAM} implementation (13 minutes) %(12min and 48secs) %768s
and our Hadoop implementation (14 minutes). %(14min and 22secs). %862s
Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
Although this additional pre-processing step is only required once for each input file, it uses an additional 13 minutes, %13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster (see Figure 1).
R and Python are also slower at pre-processing, taking 34 minutes and 92 minutes, respectively. %34min 30sec % 2070
This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
The GATK-based VCF to PED conversion for {\sc Admixture} is also slower, with a runtime of over 10 minutes.

@@

\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 30 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 86\% faster than Hadoop (29 min). %7-fold speedup
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 96\% over Python (103 min). %10 fold %25-fold
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

\section*{Materials and methods}
\subsection*{Computational resources}
We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB
memory running Ubuntu. 
For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by cloudera's CDH 5. We use Spark 1.3.1. This 13 node
cluster has a total of 416 cores and 1.22TB memory.

>>>>

As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 3 minutes. %2mins and 58secs. %178s 
This is almost 80\% faster than both the {\sc ADAM} implementation (13 minutes) %(12min and 48secs) %768s
and our Hadoop implementation (14 minutes). %(14min and 22secs). %862s
Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
Although this additional pre-processing step is only required once for each input file, it uses an additional 13 minutes, %13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster (see Figure 1).
R and Python are also slower at pre-processing, each taking approximately 34 minutes. %34min 30sec % 2070
This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
The GATK-based VCF to PED conversion for {\sc Admixture} is also slower, with a runtime of over 10 minutes.

@@

\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 30 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 86\% faster than Hadoop (29 min). %7-fold speedup
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 91\% over Python (45 min). %10 fold %25-fold
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

\section*{Materials and methods}
\subsection*{Computational resources}
We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB
memory running Ubuntu. 
For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by Cloudera's CDH 5. We use Spark 1.3.1. This 13 node
cluster has a total of 416 cores and 1.22TB memory.
